---
title: "Joins"
author: "Davin Dillon"
date: "DATA 2401"
output:
  slidy_presentation: default
  pdf_document: default
  html_document: default
---

```{r, echo = F, message = F}
library(tidyverse, warn.conflicts = F)
```

# Joins

Sometimes you store data in different dataframes. Consider: you're a charity, and you have one data frame for donations, and one for email and phone number of your donors.

- This helps with data storage: If you store every donation with the email and phone number, you're wasting space. 

- This helps with data updating: If you need to update a donor's phone number, you only have to do it in one spot, not on every row of donations they've done. 

*Those of you who have worked in SQL, this type of thinking should look familiar.* 

--- 

So, how do we do this? 

You join data through data that is in both datasets: you need a way to tell the program how to join them. 

You identify columns that are in both tables, then use these columns to "match" the corresponding rows. 

---

So, suppose we have this table of donations: 

|Name| Amount | Date |
| --- | --- | --- 
| Sam | 100 | 2/12/2019 |
| Phil | 50 | 2/12/2019 | 
| Alice | 100 | 4/23/2019 |
| Phil | 50 | 5/17/2019 |
| Maria | 100 | 6/7/2019 |
| Susanne | 50 | 7/2/2019 |

And this table of donors: 

|Name| Email |
| --- | --- 
| Sam | sam\@gmail.com 
| Alice | aliceemail\@gmail.com
| Susanne | susanneroks\@gmail.com
| Phil | phillip\@hotmail.com
| Mary | mary\@gmail.com


We would want to use "Name" to identify which rows correspond to each other in the table. 


Now, how does R decide which order they go in and what gets left in or left out? You need to specify which type of join you want. 


---

## Types of Joins

There are two main types of joins, mutating joins and filtering joins. 

- Mutating joins add all the columns from one table to the other, combining the variables. 

- Filtering joins only take the variables from the first table, but filter the rows based on whether or not that row is present in the second table. 

--- 

### Mutating Joins 

The first kind of join combines variables from both data frames. 


Straight from the R documentation: 

- `left_join(x, y, by = )`: return all rows from x, and all columns from x and y. Rows in x with no match in y will have NA values in the new columns. If there are multiple matches between x and y, all combinations of the matches are returned.

- `right_join(x, y, by = )`: return all rows from y, and all columns from x and y. Rows in y with no match in x will have NA values in the new columns. If there are multiple matches between x and y, all combinations of the matches are returned.

- `inner_join(x, y, by = )`: return all rows from x where there are matching values in y, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned.

- `full_join(x, y, by = )`:return all rows and all columns from both x and y. Where there are not matching values, returns NA for the one missing.

How are these different?

---



And again with some simple data. *This data is from the join chapter on stat545.com, linked later in the notes. *

At each step, note what variables and rows appear. 

*Note:* I don't specify which key/identifier variable to use, so R finds variables with the  same names and matches using them. (It prints out what it uses so I can make sure it guessed right.)

```{r}
# make tibbles of data manually

superheroes <- tibble::tribble(
       ~name, ~alignment,  ~gender,          ~publisher,
   "Magneto",      "bad",   "male",            "Marvel",
     "Storm",     "good", "female",            "Marvel",
  "Mystique",      "bad", "female",            "Marvel",
    "Batman",     "good",   "male",                "DC",
     "Joker",      "bad",   "male",                "DC",
  "Catwoman",      "bad", "female",                "DC",
   "Hellboy",     "good",   "male", "Dark Horse Comics"
  )


publishers <- tibble::tribble(
  ~publisher, ~yr_founded,
        "DC",       1934L,
    "Marvel",       1939L,
     "Image",       1992L
  )

superheroes
publishers
```

---

Inner join: Return all rows from x where there are matching values in y, and all columns from x and y.

What is missing? Why? 

```{r, message = F}
inner_join(superheroes, publishers)
```

Think: how will `inner_join(publishers, superheroes)` be different? 

---


```{r}
library(tidyverse)
inner_join(publishers, superheroes)
```

---

Left join: Return all rows from x, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned. 

```{r}
left_join(superheroes, publishers)
```

---

Right join: Return all rows from y, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned. 

*Note:* Not often used. This is the same as `left_join(publishers, superheroes)`

```{r}
right_join(superheroes, publishers)
```

--- 

Full join: Return all rows and all columns from both x and y. Where there are not matching values, returns NA for the one missing

```{r}
full_join(superheroes, publishers)
```


---

### Filtering Joins 

Filtering joins keep just the cases from the first data frame, but filters using the second. 

- `semi_join(x, y, by = )`: return all rows from x where there are matching values in y, keeping just columns from x.

- `anti_join(x, y, by = )`: return all rows from x where there are not matching values in y, keeping just columns from x.

---

Semi join:  Return all rows from x where there are matching values in y, keeping just columns from x. 

```{r}
semi_join(superheroes, publishers)
```

How is this different than the result from inner_join(superheroes, publishers)?

How will this compare to `semi_join(publishers, superheroes)`?

---

```{r}
semi_join(publishers, superheroes)
```

---

Anti join: Return all rows from x where there are not matching values in y, keeping just columns from x

```{r}
anti_join(superheroes, publishers)
```


I also highly suggest you read through Chapter 11 of our textbook, or Chapter 13 of R for Data Science:  https://r4ds.had.co.nz/relational-data.html
or this lecture from Jenny Bryan: https://stat545.com/join-cheatsheet.html

---

# Examples 

`band_members` and `band_instruments` are two data sets inside R that show these joins pretty well

```{R}
band_members
band_instruments
```

Looking at the two datasets above, convince yourself what each of the different joins below will give you. Once you do so, change `eval = F` to `eval = T`
```{R, eval = T}
inner_join(band_members, band_instruments)
full_join(band_members, band_instruments)
left_join(band_members, band_instruments)
right_join(band_members, band_instruments)
```

```{r, eval = F}
semi_join(band_members, band_instruments)
anti_join(band_members, band_instruments)

semi_join(band_instruments, band_members)
anti_join(band_instruments, band_members)
```


---



**Practice: **

Load the `"nycflights13"` package.  You'll also need to load `dplyr` or `tidyverse`, if you didn't above.

```{r}
library(tidyverse)
#install.packages("nycflights13")
library(nycflights13)
my_flights <- flights # store my own version
my_flights <- na.omit(my_flights) ## this dataset has data on cancelled flights too, 
                                  ## so we 'll remove those for now
head(my_flights)
View(my_flights)
View(planes)
```


Create a dataframe of the average arrival delays for each _destination_

```{r}
avg_delay_df <- my_flights %>% 
  group_by(dest) %>%
  summarize(avg_del = mean(arr_delay)) %>% 
  arrange(-avg_del)
avg_delay_df  <- rename(avg_delay_df, 'faa' = 'dest')
avg_delay_df
```

Then use `left_join()` to join on the "airports" dataframe, which has the airport information.

```{r}
adel <- left_join(avg_delay_df, airports, by = 'faa')
```


Note here: the column names don't match anymore! 

Which airport had the largest average arrival delay?

```{r}
avg_delay_df %>% 
  arrange(-avg_del)
```

```{r}
dep_del_df <- left_join(airlines, my_flights) %>% 
  group_by(name) %>% 
  summarize(avg_dep_del = mean(dep_delay)) %>% 
  arrange(avg_dep_del)
dep_del_df

```


---

Which airline had the smallest average departure delay?

To answer this, you need to create a dataframe of the average departure delay for each _airline_, then use `left_join()` to join on the "airlines" dataframe

 - Create a dataframe of the average departure delay for each _airline_
 - use `left_join()` to join on the "airlines" dataframe
 - use slice_???() to find the result you need




```{r}

View(dep_del_df)
left_join(airlines, dep_del_df) %>% 
  arrange(avg_dep_del)
```


Examples from R4DS:    

```{r}

top_dest <- flights %>%
  count(dest, sort = TRUE) %>%
  head(10)
top_dest
View(airlines)
```

Now you want to find each flight that went to one of those destinations. You could, of course, filter. But! Now you can semi-join: 

```{r}
flights %>% 
  semi_join(top_dest)
```


What does anti_join(flights, airports, by = c("dest" = "faa")) tell you? What does anti_join(airports, flights, by = c("faa" = "dest")) tell you?

Anti_join(flights, airports, by = c("dest" = "faa")) should give us the flights that do not have a destination and anti_join(airports, flights, by = c("faa" = "dest")) should give us the airports that do not have flights coming in to them.



